Spatiotemporal Multi-Resolution Approximation of the Amari Type Neural Field Model

### 2012-05-14 Multiresolution_framework/ltx/manuscript.tex
added Viktor comments

Neural field models describe the mass action of the central nervous system and are a critical link in our understanding of the biophysics of the EEG. The key components of these models typically describe the macroscopic dynamics of the human brain, but can also descriptive of finer-scale neurodynamics. Examples are the post-synaptic response kernels (the pulse to wave function) and the activation function (wave to pulse function) \citep{Jirsa1997}. The former can be descriptive of inhibitory and excitatory post-synaptic potentials in microscopic single neural models, such as integrate and fire, or an ensemble response to stimulation in macroscopic neural mass models. The latter, can be used to model spiking statistics across spatial scales from single neurons (where it can be considered a CDF of the probability of firing) to mean firing rates of neural masses. The key factor linking the key functions of the model to particular spatial scales is geometry of the connectivity, hence the ability to infer a multi-resolution connectivity kernel will enable to the formulation of a multi-resolution neural field model.

The utility of these models for answering questions in clinical neuroscience and neurology lies in the ability to infer parameters from electrophysiological data. The ability to create patient-specific neural models will contribute to our knowledge base of diseases such as epilepsy and will enable the development of new treatment strategies. This is particularly relevant to the advent of new devices that use therapeutic electrical stimulation. Stimulation strategies for devices currently operate in an open loop, where stimulation parameters are chosen using a process of trial and error. Therefore, there exists enormous potential to improve the performance of these devices using systems theory, which in turn requires a suitable dynamic model. Neural mass and neural field models are ideal candidates for estimation algorithms due to their strong links with physiology and their parsimony. It is expected that parameters of the neural fields models will be patient-specific and will, therefore, need to be inferred from data.

@@

The multi-resolution structure has been described by numerous methods. For example, this idea of multi-resolution can be thought of a generalization of modeling the layers of the cortex separately, where the connectivity kernel, or neural foot-print, has varying widths \citep{Wilson1973}.

Typically, the experimentalist only has access the net field generated by the multi-resolution architecture of the cortex. Follow this, it is natural to ask how do we best describe the multi-resolution structure of the neural system. The solution to this problem was proposed by \citet{Breakspear2005} in using a wavelet decomposition. This paper is directed at developing a framework for patient-specific multi-resolution neural field models, forming a fundamental link between data and theory.

@@

The single layer homogeneous neural field model, in the most general form, is described by the equation
\begin{equation}
	\label{FullDoubleIntModel} \underbrace{v\left(\mathbf{r},t\right)}_{\text{field}} =
	\int_{-\infty}^t 
	\underbrace{h\left(t - t'\right)}_{\text{synapse}} \int_\Omega
	\parham{\theta}\underbrace{w\left(\mathbf{r},\mathbf{r}'\right)}_{\text{connectivity}} 
	\underbrace{f\left( v\left( \mathbf{r}',t' \right)\right)}_{\text{firing}}
	\, \mathrm{d}\mathbf{r}'\mathrm{d}t',
\end{equation}
where the post-synaptic membrane voltage at time $t$ of a population of neurons at position $\mathbf r$ within the spatial domain $\Omega$ is denoted by $v\left(\mathbf r,t\right)$. The function $h(t)$ describes the post-synaptic response kernel, $w\left(\mathbf{r},\mathbf{r}'\right)$ describes the connectivity kernel and $f(v(\mathbf{r},t))$ describes the relationship between the mean membrane potential and the mean firing rate. The parameter $\theta$ can be thought of the synaptic efficiency or coupling strength, where $\theta>0$ and $\theta<0$ describes excitation and inhibition, respectively. There are various forms of $h(t)$ that have been proposed in the literature, with different levels of complexity. These are shown in Fig.~\ref{fig:Modelcomponents}. Similarly, both the connectivity kernel and the activity function (firing rate) can take various forms. These are also shown in Fig.~\ref{fig:Modelcomponents}. The form of $w(\mathbf{r},\mathbf{r}')$ typically describes an Laplacian (exponential) or Gaussian decay in connectivity strength from any point in the field \cite{Breakspear2010,Atay2005}. By assuming the form of $h(t)$ is the same across layers (scales) the Amari style neural field model is formed, incorporating a multilayer structure into the connectivity kernel. In this way, the connectivity kernel typically takes the form of a Mexican hat (Gaussian shape within layer connectivity) or Wizard hat (Laplacian shape within layer) as depicted in Fig.~\ref{fig:Modelcomponents}.

The stochastic IDE form of the Amari neural field formulation~\cite{Amari1977} is given by (see~\cite{Freestone2011} for a full derivation)
\begin{equation}\label{eq:DiscreteTimeModel}
	v_{t+1}\left(\mathbf{r}\right) = 
	\xi v_t\left(\mathbf{r}\right) + 
	T_s \int_\Omega { 
	    w\left(\mathbf{r},\mathbf{r'}\right)
	    f\left(v_t\left(\mathbf{r}'\right)\right) 
	\, \mathrm{d}\mathbf{r}'} 
	+ e_t\left(\mathbf{r}\right), 
\end{equation}
where in this first-order (in time) form of the model the membrane dynamics are included through the parameter $\xi=1-T_s/\tau$, $\tau$ is the membrane time constant and $T_s$ is the sampling time step. 
% The connectivity strength between neural populations at a distance $\mid\mathbf{r}-\mathbf{r'}\mid$ is described by the connectivity kernel $w\left(\mathbf{r}-\mathbf{r}'\right)$. The connectivity kernel is taken as a ``Mexican hat'' function, which describes local excitation and lateral inhibition \cite{Amari1977,Atay2005}. The components of the model are shown in Fig.~\ref{fig:Modelcomponents}
\begin{figure}[!t]
\centering
\includegraphics[scale=1]{./Graph/fig2.pdf}
\caption{ {\bf The IDE neural field model components}. (\textbf{A}) The Mexican hat (dotted line) and Wizard hat (solid line) connectivity kernels. (\textbf{B}) First order (dotted line) and second order (solid line) synaptic kernels. (\textbf{C}) A representative of the activation function.}
\label{fig:Modelcomponents}
\end{figure}
The term $e_t(\mathbf r)$ is a zero mean Gaussian disturbance, spatially colored but temporally white, with covariance function 
\begin{equation}
cov\left(e_{t}\left(\mathbf{r}\right),e_{t+t'}\left(\mathbf{r'}\right)\right)=\sigma_d\eta(\mathbf{r}-\mathbf{r'})\delta(t-t'),
\label{eq:FieldDisturbance}
\end{equation}
where \parham{$\eta(\mathbf{r}-\mathbf{r'})$ is a spatially homogeneous covariance function, $\sigma_d$ is disturbance variance} and $\delta(\cdot)$ is the Dirac-delta function. %$ = \varsigma v_t(\mathbf{r})$ \cite{VanRotterdam1982,Murphy2009}.

@@

It is important to point out there is no doubt that there will always be a mismatch between the model and actual cortical tissue. Nevertheless, the simplifications used in formulating the equations enable the development of new methods for extracting information from electrophysiological data, that has the potential to influence the treatment of diseases such as epilepsy. The trade-off between parsimony and complexity is always difficult when considering neurodynamics and this is further complicated when using estimation algorithms that have huge computational demands with such high-dimensional systems.

@@

The EM algorithm was allowed to run until a maximum number of 20 iterations was reached, though typically the change in the transition matrix Frobenius norm dropped below $10^{-6}$ after less than 10 iterations. The rate of convergence of the EM based algorithm is depicted in Fig.~\ref{fig:MRA-Convergence}. 
\begin{figure}[t]
 	\centering
 		\includegraphics[scale=1]{./Graph/fig8.pdf}
 		\caption{{\bf Convergence of the EM based algorithm}. Representative plot of
the $\mathbf A(\boldsymbol\theta)$ Frobenius norm vs iterations of EM based algorithm. The change in the stopping criterion falls below $10^{-6}$ after less than 10 iterations.} 
\label{fig:MRA-Convergence}  
 \end{figure}     
% The spatial cutoff frequency of the observed field was used to specify the level of decomposition required to represent the field using the wavelets. Using an oversampling parameter of 5, the cutoff spatial frequency was $\nu_{cy} \approx 6.9 $ cycles/mm, obtained by averaging the spectral power of the spatial frequency (over time) from the observations \cite{Scerri2009}. Therefore, wavelets up to level $j=3$ (with the bandwidth $\approx[5,8]$ cycles/mm) can represent the significant spatial characteristics of the field, yielding $n_x = 131$ states. 
The actual connectivity kernel and the decomposition is plotted in Fig.~\ref{fig:KernelEstimate}(A). No assumptions where made about the shape of the kernel. The reconstructed kernel is in good accordance with the actual kernel, where the actual kernel lies inside the confidence interval. The large standard deviation is due to the high number of parameters need to be estimated. The small error in the estimate is attributed to the MRA of the system used to form the estimator. Figs.~\ref{fig:KernelEstimate}(B) and (C) illustrate the kernel scaling and wavelet functions, respectively. A total number of 25 basis functions were used to reconstruct the kernel, comprising 13 and 12 scaling and wavelet functions at $j=1$, respectively. 
\begin{figure}[!t]  
 \centering
\includegraphics{./Graph/fig9.pdf}
\caption{ {\bf Connectivity kernel estimate}. (\textbf{A}) The actual kernel and its components are shown by solid and dashed black lines, respectively. The mean kernel estimates over 200 realizations and confidence interval are shown by the red line and red shaded region ($\pm2$ std.). (\textbf{B},\textbf{C}) Kernel scaling and wavelets functions, respectively, at $j=1$ used in the estimation algorithm.}
\label{fig:KernelEstimate}
\end{figure}

@@

In this paper, we have presented a novel model-based framework for estimating cortical dynamics from electrophysiological measurements. The novel and key developments of the paper include the multi-resolution finite element model representation of the neural field, and the estimator for an intracortical connectivity kernel with an arbitrary shape. This work is significant, as the ability to create patient specific neural field models has the potential to contribute to our understanding and improve treatment of diseases resulting from abnormal neurodynamics, such as epilepsy. Other groups have previously highlighted the importance of a multi-resolution approach in neural field modeling. It is thought that the dynamics and connectivity structure differs at different spatial scales~\citep{Qubbaj2009,Breakspear2005,Schultze-Kraft2011}. %\dean{This is a test to see if we can fit something about Breaky's paper in here. It would be nice to slot something in here because he used wavelets.} 

@@

Additional extension for future work include extending the model to capture a second order synaptic response kernel, The second-order postsynaptic kernel proposed in \citet{VanRotterdam1982} leads to dynamics that are more reminiscent to real postsynaptic potentials. Another extension of this framework is the inclusion of finite distant-dependent action potential propagation velocity into the parametric neural field equations. In small networks the delays via propagation can be neglected. However, delays must be accounted for in large scale networks. For instance, it is shown by \citet{Ghosh2008} that inclusion of time delays arising from propagation along connecting fibres are essential for the fluctuations to be appeared in the default networks. Finally, future work should be directed towards applying and validating the framework on real data.

@@

Twice the logarithm of the above distribution, once expanded, is given by

@@

The factor of 2 is included to simplify the derivation in the later steps. Substituting $\mathbf A( \boldsymbol\theta)$ from Eq.~\eqref{eq:A_theta} into Eq.~\eqref{eq:Qfunction} gives

>>>>

Neural field models describe the mass action of the central nervous system and are a critical link in our understanding of the biophysics leading to the generation of  Electroencephalography (EEG) and Magnetoencephalography (MEG) signals. The key components of these models typically describe the macroscopic dynamics of the human brain, but can also descriptive of finer-scale neurodynamics. Examples are the post-synaptic response kernels (the pulse to wave function) and the activation function (wave to pulse function) \citep{Jirsa1996,Jirsa1997}. The former can be descriptive of inhibitory and excitatory post-synaptic potentials in microscopic single neural models, such as integrate and fire, or an ensemble response to stimulation in macroscopic neural mass models \citep{Stefanescu2008}. The latter, can be used to model spiking statistics across spatial scales from single neurons (where it can be considered a cumulative distribution function (CDF) of the probability of firing) to mean firing rates of neural masses. The key factor linking the key functions of the model to particular spatial scales is geometry of the connectivity, hence the ability to infer a multi-resolution connectivity kernel will enable to the formulation of a multi-resolution neural field model.

The utility of these models for answering questions in clinical neuroscience and neurology lies in the ability to infer parameters from electrophysiological data. The ability to create patient-specific neural models will contribute to our knowledge base of diseases such as epilepsy and will enable the development of new treatment strategies. Large-scale neuroinformatic efforts such as the Virtual Brain project (\href{http://www.thevirtualbrain.org}{www.thevirtualbrain.org} \citep{Jirsa2010}) are on the way to enable patient-specific full brain modeling. This is particularly relevant to the advent of new devices that use therapeutic electrical stimulation. Stimulation strategies for devices currently operate in an open loop, where stimulation parameters are chosen using a process of trial and error. Therefore, there exists enormous potential to improve the performance of these devices using systems theory, which in turn requires a suitable dynamic model. Neural mass and neural field models are ideal candidates for estimation algorithms due to their strong links with physiology and their parsimony. It is expected that parameters of the neural fields models will be patient-specific and will, therefore, need to be inferred from data.

@@

The multi-resolution structure has been described by numerous methods. For example, this idea of multi-resolution can be thought of as the generalization of modeling the layers of the cortex separately, where the connectivity kernel, or neural foot-print, has varying widths \citep{Wilson1973}.

Typically, the experimentalist only has access to the net field generated by the multi-resolution architecture of the cortex. Following this, it is natural to ask how do we best describe the multi-resolution structure of the neural system. The solution to this problem was proposed by \citet{Breakspear2005} in using a wavelet decomposition. This paper is directed at developing a framework for patient-specific multi-resolution neural field models, forming a fundamental link between data and theory.

@@

The single layer homogeneous neural field model, in the most general form, is described by the equation
\begin{equation}
	\label{FullDoubleIntModel} \underbrace{v\left(\mathbf{r},t\right)}_{\text{field}} =
	\int_{-\infty}^t 
	\underbrace{h\left(t - t'\right)}_{\text{synapse}} \int_\Omega
	\parham{\theta}\underbrace{w\left(\mathbf{r},\mathbf{r}'\right)}_{\text{connectivity}} 
	\underbrace{f\left( v\left( \mathbf{r}',t' \right)\right)}_{\text{firing}}
	\, \mathrm{d}\mathbf{r}'\mathrm{d}t',
\end{equation}
where the post-synaptic membrane voltage at time $t$ of a population of neurons at position $\mathbf r$ within the spatial domain $\Omega$ is denoted by $v\left(\mathbf r,t\right)$. The function $h(t)$ describes the post-synaptic response kernel, $w\left(\mathbf{r},\mathbf{r}'\right)$ describes the connectivity kernel and $f(v(\mathbf{r},t))$ describes the relationship between the mean membrane potential and the mean firing rate. The parameter $\theta$ can be thought of as the synaptic efficiency or coupling strength, where $\theta>0$ and $\theta<0$ describe excitation and inhibition, respectively. There are various forms of $h(t)$ that have been proposed in the literature, with different levels of complexity. These are shown in Fig.~\ref{fig:Modelcomponents}. Similarly, both the connectivity kernel and the activity function (firing rate) can take various forms. These are also shown in Fig.~\ref{fig:Modelcomponents}. The form of $w(\mathbf{r},\mathbf{r}')$ typically describes an Laplacian (exponential) or Gaussian decay in connectivity strength from any point in the field \cite{Breakspear2010,Atay2005}. By assuming the form of $h(t)$ is the same across layers (scales) the Amari style neural field model is formed, incorporating a multilayer structure into the connectivity kernel. In this way, the connectivity kernel typically takes the form of a Mexican hat (Gaussian shape within layer connectivity) or Wizard hat (Laplacian shape within layer) as depicted in Fig.~\ref{fig:Modelcomponents}.

The stochastic IDE form of the Amari neural field formulation~\cite{Amari1977} is given by (see~\cite{Freestone2011} for a full derivation)
\begin{equation}\label{eq:DiscreteTimeModel}
	v_{t+1}\left(\mathbf{r}\right) = 
	\xi v_t\left(\mathbf{r}\right) + 
	T_s \int_\Omega { 
	    w\left(\mathbf{r},\mathbf{r'}\right)
	    f\left(v_t\left(\mathbf{r}'\right)\right) 
	\, \mathrm{d}\mathbf{r}'} 
	+ e_t\left(\mathbf{r}\right), 
\end{equation}
where in this first-order (in time) formulation of the model the membrane dynamics are included through the parameter $\xi=1-T_s/\tau$, $\tau$ is the membrane time constant and $T_s$ is the sampling time step. 
% The connectivity strength between neural populations at a distance $\mid\mathbf{r}-\mathbf{r'}\mid$ is described by the connectivity kernel $w\left(\mathbf{r}-\mathbf{r}'\right)$. The connectivity kernel is taken as a ``Mexican hat'' function, which describes local excitation and lateral inhibition \cite{Amari1977,Atay2005}. The components of the model are shown in Fig.~\ref{fig:Modelcomponents}
\begin{figure}[!t]
\centering
\includegraphics[scale=1]{./Graph/fig2.pdf}
\caption{ {\bf The IDE neural field model components}. (\textbf{A}) The Mexican hat (dotted line) and Wizard hat (solid line) connectivity kernels. (\textbf{B}) First order (dotted line) and second order (solid line) synaptic kernels. (\textbf{C}) A representative of the activation function.}
\label{fig:Modelcomponents}
\end{figure}
The term $e_t(\mathbf r)$ is a zero mean Gaussian disturbance, spatially colored but temporally white, with covariance function 
\begin{equation}
cov\left(e_{t}\left(\mathbf{r}\right),e_{t+t'}\left(\mathbf{r'}\right)\right)=\sigma_d\eta(\mathbf{r}-\mathbf{r'})\delta(t-t'),
\label{eq:FieldDisturbance}
\end{equation}
where \parham{$\eta(\mathbf{r}-\mathbf{r'})$ is a spatially homogeneous covariance function, $\sigma_d$ is disturbance variance} and $\delta(\cdot)$ is the Dirac-delta function. %$ = \varsigma v_t(\mathbf{r})$ \cite{VanRotterdam1982,Murphy2009}.

@@

It is important to point out there is no doubt that there will always be a mismatch between the model and actual cortical tissue. Nevertheless, the simplifications used in formulating the equations enable the development of new methods for extracting information from electrophysiological data. The trade-off between parsimony and complexity is always difficult when considering neurodynamics and this is further complicated when using estimation algorithms that have huge computational demands with such high-dimensional systems.

@@

The EM algorithm was allowed to run until a maximum number of 20 iterations was reached, though typically the change in the transition matrix Frobenius norm dropped below $10^{-6}$ after less than 10 iterations. The rate of convergence of the EM based algorithm is depicted in Fig.~\ref{fig:MRA-Convergence}. 
\begin{figure}[t]
 	\centering
 		\includegraphics[scale=1]{./Graph/fig8.pdf}
 		\caption{{\bf Convergence of the EM based algorithm}. Representative plot of
the $\mathbf A(\boldsymbol\theta)$ Frobenius norm vs iterations of EM based algorithm. The change in the stopping criterion falls below $10^{-6}$ after less than 10 iterations.} 
\label{fig:MRA-Convergence}  
 \end{figure}     
% The spatial cutoff frequency of the observed field was used to specify the level of decomposition required to represent the field using the wavelets. Using an oversampling parameter of 5, the cutoff spatial frequency was $\nu_{cy} \approx 6.9 $ cycles/mm, obtained by averaging the spectral power of the spatial frequency (over time) from the observations \cite{Scerri2009}. Therefore, wavelets up to level $j=3$ (with the bandwidth $\approx[5,8]$ cycles/mm) can represent the significant spatial characteristics of the field, yielding $n_x = 131$ states. 
The actual connectivity kernel and the decomposition is plotted in Fig.~\ref{fig:KernelEstimate}(A). No assumptions where made about the shape of the kernel. The reconstructed kernel is in good accordance with the actual kernel, where the actual kernel lies inside the confidence interval. The large standard deviation is due to the high number of parameters needed to be estimated. The small error in the estimate is attributed to the MRA of the system used to form the estimator. Figs.~\ref{fig:KernelEstimate}(B) and (C) illustrate the kernel scaling and wavelet functions, respectively. A total number of 25 basis functions were used to reconstruct the kernel, comprising 13 and 12 scaling and wavelet functions at $j=1$, respectively. 
\begin{figure}[!t]  
 \centering
\includegraphics{./Graph/fig9.pdf}
\caption{ {\bf Connectivity kernel estimate}. (\textbf{A}) The actual kernel and its components are shown by solid and dashed black lines, respectively. The mean kernel estimates over 200 realizations and confidence interval are shown by the red line and red shaded region ($\pm2$ std.). (\textbf{B},\textbf{C}) Kernel scaling and wavelets functions, respectively, at $j=1$ used in the estimation algorithm.}
\label{fig:KernelEstimate}
\end{figure}

@@

In this paper, we have presented a novel model-based framework for estimating cortical dynamics from electrophysiological measurements. The novel and key developments of the paper include the multi-resolution finite element model representation of the neural field, and the estimator for an intracortical connectivity kernel with an arbitrary shape. This work is significant, as the ability to create patient-specific neural field models has the potential to contribute to our understanding and improve treatment of diseases resulting from abnormal neurodynamics, such as epilepsy. Other groups have previously highlighted the importance of a multi-resolution approach in neural field modeling, where the dynamics and connectivity structure differs at different spatial scales~\citep{Jirsa2002,Jirsa2004,Qubbaj2009,Breakspear2005,Schultze-Kraft2011}.

@@

Additional extension for future work include extending the model to capture a second order synaptic response kernel, The second-order postsynaptic kernel proposed in \citet{VanRotterdam1982} leads to dynamics that are more reminiscent to real postsynaptic potentials. Another extension of this framework is the inclusion of finite distant-dependent action potential propagation velocity into the parametric neural field equations. In small networks the delays via propagation can be neglected. However, delays must be accounted for in large scale networks. For instance, it is shown by \citet{Ghosh2008} that inclusion of time delays arising from propagation along connecting fibres are essential for the fluctuations in resting state networks (see \citet{Deco2010} for review). Finally, future work should be directed towards applying and validating the framework on real data.

@@

The logarithm of the above distribution, once expanded, is given by

@@

The factor of 2 in front of the logarithm is included to simplify the derivation in the later steps. Substituting $\mathbf A( \boldsymbol\theta)$ from Eq.~\eqref{eq:A_theta} into Eq.~\eqref{eq:Qfunction} gives
